---
title: "R ventures"
author: "Sarah Hu"
date: "10/2/2021"
output: 
  html_document:
    number_sections: true
    theme: lumen
    highlight: kate
    collapsed: false
    toc: true
    toc_depth: 4
    toc_float: true
    
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```
# Introduction

The sections below are a compilation of recommended resources in R and common data wrangling practices I use in R that I've 'tutorial-ized'. I welcome additions and recommendations! Note that if it is written below, I probably use it often in my own data analysis.

First, make sure you're familiar with your R set up. I typically code in R Studio and there are already a ton of great resources for installing and navigating R and R Studio. Everyone's set up may be a little bit different - I've compiled some resources and notes on each resource that might help you decide on how to proceed.

## Installing R & RStudio

### Tutorials on R installation & start-up
If you're brand new to R and the command line, explore the resources through Software Carpentry. This is a great resource for learning how to install RStudio on your computer and getting oriented in the R interface - [Introduction to R and RStudio](https://swcarpentry.github.io/r-novice-gapminder/).

Through the [BVCN](https://biovcnet.github.io/) we offered an introduction to R using the browser-supported RStudio cloud. If you already have RStudio installed and working, you can follow [Lesson 1](https://github.com/biovcnet/biovcnet.github.io/wiki/TOPIC%3A-R). We even have a [video](https://www.youtube.com/watch?v=u6vgWyD351g&list=PL4K-daRUS2A9noShSQjLfFGnaXg7YelrO&index=2).

Another fantastic resource is [Riffomonas](https://riffomonas.org/minimalR/). See Pat Schloss's instructions for [installing R](https://riffomonas.org/minimalR/00_installation.html). I prefer video content when learning new skills, so I appreciate the work Pat's put into his website and teaching mini R tutorials. 

### R via Anaconda
If your life is maintained through Anaconda environments, you may want to check out [my post on using conda and R](https://alexanderlabwhoi.github.io/post/anaconda-r-sarah/). This is how I run R on my computer and it works for me, but it is so I can maintain several R versions and separate environments for different project.

# Importing data

If you are unfamiliar with the command line and want to learn more about paths and how to navigate in your terminal I'd recommend following [this crash course in Unix](https://github.com/biovcnet/biovcnet.github.io/wiki/TOPIC%3A-Unix). The below instructions will assume you have an understanding of how R is installed on your computer and you can locate the files you need!

_Note_ that gaining a little bit of experience in Unix and command line efficiency is recommended because this is the same mode of thinking required to build up your R environment and navigate to access various files. However, the same commands that work on the command line are not in the R programming language. While many of the words and terms are similar, when you launch R - you're entering an environment where the R language is required instead. 
## Data types that work best

One of the more frustrating components of learning any computer language or program is _figuring out what I need to start_. The most common issues come from a lot of hidden characters and weird inconsistencies between computers and software. Two main recommendations:
* Download and get familiar with a good text editor. I use *Text Wrangler* / *BBEdit* [get it here](https://www.barebones.com/products/bbedit/faqs.html).
* Get familiar with _.csv_ file types. The way data is separated (is it a tab or a comma or a space?) in a data frame is what you need to be aware of when importing data into R. .csv files are a standard file type, so if you're sharing your data with a collaborator or friend, this is the most straight-forward way to do it to avoid any incompatibilities.


## Examples of importing data

Grab your own .csv file or download one here. *To download*, right click

```{r}
asv_18s_raw <- read.csv(file = "data-input/asv-gordaridge-18s-wide.csv")
```


## Data management & organization

I'd recommend you work with [RStudio projects](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects). By containing a project within an RStudio Project, I simply open the *.Rproj* and it opens up RStudio to the location I need to access all the relevant code and data files to work on my project. Software Carpentry also maintains a good [how to use RStudio projects](https://swcarpentry.github.io/r-novice-gapminder/02-project-intro/index.html) resource that I recommend. 

For each project I work on, I maintain a */data-input* folder (or directory) that has all of my files I need to import for analysis. Similarly, I typically make a separate directory for output files and figures.

_I'd recommend_ you make up from rules for yourself. My main coding advice is always: _keep current you organized, so future you thanks past you_. Add lots of notes for yourself on how you organize your work and how you leave it! Then when you open it back up and realize you've forgotten everything - there is a nice note from past you. 

As of now (mid-2021), I have one directory on my computer per project. Within that directory I have one RStudio project and directories for raw data, input and processed data, and output folders for products. My own rule is that I can re-create the entire directory file structure and outputs by running the R code with all the raw input data. 

## Get data into R

## Common _first steps_ to take in my newly imported data

# Get basic stats from data

# Data wrangling

# Plotting my data

# Practical Ecology in R
